---
title: "Case_Study_1_Analysis"
author: "Team 100"
date: "1/15/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load}
library(BMS)
library(dplyr)
library(tidyr)
Longnecker <- readRDS("/cloud/project/Longnecker.rds")
```


## Correlation

**This may be a little misleading because I converted everything into numeric (including smoking status, centers, and race). I'll figure out factors later for these I hope.

```{r}
library(corrplot)
corrplot.mixed(cor(Longnecker %>% select(starts_with("pcb_")) %>% drop_na() %>% sapply(.,as.numeric)), order="hclust", tl.col="black")
corrplot.mixed(cor(Longnecker %>% select(-starts_with("pcb_"),-albumin) %>% drop_na() %>% sapply(.,as.numeric)), order="hclust", tl.col="black")
```

We observe that PCB variations are positively correlated with one another, and that certain groups of variables are also correlated (education, occupation, & income; triglycerides & cholesterol; race & center; race & dde; maternal age & triglycerides, etc.).

## Variable Selection

We will use lm and BMA to explore variable selection. By getting rid of the albumin and score columns, we have very little missing data (2379/2380 original observations).

```{r}
#removing uninterpretable (at the moment) vars and albumin
Longnecker_potential<-Longnecker %>% 
  select(-albumin,-score_education,-score_income,-score_occupation)

#sum PCBs, treat race and center as factor random effects 
#if someone knows a better way to pivot_wide race and center lmk because this was a pain
Longnecker_potential <- Longnecker_potential %>%
  mutate(pcb= Longnecker_potential %>% 
           select(starts_with("pcb")) %>% rowSums()) %>%
  select(-starts_with("pcb_")) %>%
  pivot_wider(names_from=race,values_from=race) %>%
  mutate(white=as.numeric(white=="white"),
         black=as.numeric(black=="black"),
         other=as.numeric(other=="other")) %>%
  replace_na(list(white=0,black=0,other=0)) %>%
  pivot_wider(names_from=center,values_from=center,values_fill = list(center=0),values_fn = list(center=function(x) x/x)) 

#moving gest_age to the front of the data frame
Longnecker_potential<-cbind(Longnecker_potential %>% select(gestational_age),Longnecker_potential %>% select(-gestational_age))
names(Longnecker_potential)[11:22]<-paste(LETTERS[1:12],LETTERS[1:12],sep="_")


#simple lm model suggests race, center, and triglycerides have significant effects
#terrible R-squared, but this might be taken care of using transformations and splitting by our random effects
lm_model<-lm(gestational_age ~ .,Longnecker_potential)
summary(lm_model)

#BMA model 
mfls <- bms(Longnecker_potential, burn=100000, iter=200000, g="BRIC", mprior="uniform", nmodel=2000, mcmc="bd", user.int=FALSE)
coef(mfls,exact=TRUE)
```

The results from BMA suggest that triglycerides, centers, and race have non-zero coefficients for most of the models explored. DDE is also non-zero in a significant proportion of them. This is corroborated by the naive linear model. 

```{r}
bms(gestational_age ~ dde+triglycerides+black+E_E+F_F, burn = 1000, iter = NA, nmodel = 1000, mcmc = "bd",
  g = "UIP", mprior = "uniform", mprior.size = NA, user.int = TRUE, 
    start.value = NA, g.stats = TRUE, 
    logfile = FALSE, logstep = 10000, force.full.ols = FALSE, 
    fixed.reg = numeric(0), data = Longnecker_potential, randomizeTimer = TRUE)
```

